Multi-Domain Adaptation in Neural Machine Translation with dynamic sampling strategy

### 2021-05-07 2021/emnlp/auto_cl_mdmt.tex
change title

\title{Automated curriculum learning for Multi-Domain NMT by using loss-driven reward}

@@

Several recent proposals \citep{platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Graves17automated,Wang20balancing,Wang20learning-multi,Vanderwees17dynamic} have explored ways to instead consider \emph{dynamic} data selection and sampling strategies that surpass static strategies. Furthermore \citet{Zhang19curriculum,Vanderwees17dynamic} construct a fixed curriculum while \citet{platanios19competence,Kumar19reinforcement,Graves17automated,Wang20balancing,Wang20learning-multi} build curricula automatically adapted to the training.

@@

Our method MDAC constructs an automatically adapting distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a fixed data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution via REINFORCE algorithm \citep{Williams92simple}.

\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * \displaystyle{\mathop{\sum}_{d=1}^K} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi} \\
\end{align*}
\begingroup
\allowdisplaybreaks
With the following reward computation:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\lambda^t) - J^t(\theta_t,\lambda^t)
\end{align*}
Where 
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& \theta_{t+i-1} - \mathbf{lr}_{nmt} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta} \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^K}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d)
\end{array}
\end{equation}
\endgroup
$\mathbf{lr}_{data}$ is the learning rate of the parameter of the sampler. $\mathbf{lr}_{nmt}$ is the learning rate of the parameter of the NMT system. $J^t(\theta,\lambda_t)$ is the weighted loss computed over $K$ dev-sets corresponding to $K$ target domains.

@@

The work of \cite{Wang20balancing} based on the Bi-level Optimization framework, which aims to find an optimal fixed distribution $\lambda^{l}$ that will result in the best NMT model regarding a given target dev set at the end of the training, derives also a similar form of update for $\psi$. However, their reward is the cosine similarity between the gradient computed with the training data and the gradient computed with the dev set.

@@

\item Our comparison of multi-domain systems includes baselines with fixed data mixtures corresponding to $\vlambda_0$ and $\vlambda_1$, and our reimplementations of recent proposals from the literature: Curriculum Learning of \cite{Zhang19curriculum} and Differential Data Selection of \cite{Wang20balancing}.
\end{itemize}

@@

MDAC outperforms almost fixed training distribution including $\big[ \lambda_0, \lambda_{0.75}, \lambda_{1.0}\big]$ with significant margin while performs slightly better than $\big[ \lambda_{0.25}, \lambda_{0.5} \big]$. This observation indicates that MDAC allows us to skip the heuristic choice for the training mixture of different domains or tasks.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    \multicolumn{9}{|c|}{\sl Supervised domain adaptation} \\ \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \system{MDAC($\vlambda_1, \vlambda_2$)}&36.2&-&-&-&46.1&-&-&- \\
    \system{DDS($\vlambda_1, \vlambda_2$)}&36.5&-&-&-&44.4&-&-&- \\
    \system{MDAC($\vlambda_1, \vlambda_2$)}&-&57.2&53.8&-&-&-&-&- \\
    \system{DDS($\vlambda_1, \vlambda_2$)}&-&56.1&52.6&-&-&-&-&- \\
    \system{MDAC($\vlambda_1, \vlambda_2$)}&-&-&-&32&-&90&-&- \\
    \system{DDS($\vlambda_1, \vlambda_2$)}&-&-&-&32.9&-&91.1&-&- \\ \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, all the non empty columns \emph{correspond to the same bi-domain system}}
  \label{tab:bi-da}
\end{table*}

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    \multicolumn{9}{|c|}{\sl Supervised domain adaptation} \\ \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \multicolumn{9}{|c|}{\sl Unsupervised domain adaptation} \\ \hline
    \system{Mixed-0}      & - & - & - & - & - & - & - & 22.4 \\
    \system{Mixed-0.25} & - & - & - & - & - & - & - & 22.7\\
    \system{Mixed-0.5}   & - & - & - & - & - & - & - & 23.6\\
    \system{Mixed-0.75} & - & - & - & - & - & - & - & 23.7\\
    \system{Mixed-1} & - & - & - & - & - & - & - & 23.5\\
    \hline \hline
    \system{MDAC($\vlambda_0, \vlambda_{news}$)} &-&-&-&-&-&-&-&23.1 \\
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &-&-&-&-&-&-&-&23.3 \\ \hline\hline 
    \multicolumn{9}{|c|}{\sl 30 clusters} \\ \hline
    \system{MDAC($\vlambda^*, \vlambda_d$)}&36.3&56.9&51.3&33.2&43.4&89.4&51.8&-\\
     \system{DDS($\vlambda^*, \vlambda_d$)}&35.3&55.2&49.4&30.6&43.3&89.9&50.6&-\\ \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, each non empty column correspond to one expert system.}
  \label{tab:unsupervised-da}
\end{table*}

>>>>

\title{Multi-Domain Adaptation in Neural Machine Translation with dynamic sampling strategy}

@@

Several recent proposals \citep{platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Graves17automated,Wang20balancing,Wang20learning-multi,Vanderwees17dynamic} have explored ways to instead consider \emph{dynamic} data selection and sampling strategies that surpass static strategies. Furthermore \citet{Zhang19curriculum,Vanderwees17dynamic} construct a static curriculum while \citet{platanios19competence,Kumar19reinforcement,Graves17automated,Wang20balancing,Wang20learning-multi} build curricula automatically adapted to the training.

@@

Our method MDAC constructs an automatically adapting distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution via REINFORCE algorithm \citep{Williams92simple}.

\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{data} * \displaystyle{\mathop{\sum}_{d=1}^K} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi} \\
\end{align*}
\begingroup
\allowdisplaybreaks
With the following reward computation:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t)
\end{align*}
Where 
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& \theta_{t+i-1} - \mathbf{lr}_{nmt} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta} \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d)
\end{array}
\end{equation}
\endgroup
$\mathbf{lr}_{data}$ is the learning rate of the parameter of the sampler. $\mathbf{lr}_{nmt}$ is the learning rate of the parameter of the NMT system. $J^t(\theta,\lambda_t)$ is the weighted loss computed over $K$ dev-sets corresponding to $K$ target domains.

@@

The work of \cite{Wang20balancing} based on the Bi-level Optimization framework, which aims to find an optimal static distribution $\lambda^{l}$ that will result in the best NMT model regarding a given target dev set at the end of the training, derives also a similar form of update for $\psi$. However, their reward is the cosine similarity between the gradient computed with the training data and the gradient computed with the dev set.

@@

\item Our comparison of multi-domain systems includes baselines with fixed data mixtures corresponding to $\lambda^l \in \big[ \lambda_0, \lambda_{0.25}, \lambda_{0.5}, \lambda_{0.75}, \lambda_{1.0}\big]$, and our reimplementations of recent proposals from the literature: Curriculum Learning of \cite{Zhang19curriculum} and Differential Data Selection of \cite{Wang20balancing}.
\end{itemize}

@@

MDAC outperforms static training distributions including $\big[ \lambda_0, \lambda_{0.75}, \lambda_{1.0}\big]$ with significant margin while performs slightly better than $\big[ \lambda_{0.25}, \lambda_{0.5} \big]$. This observation indicates that MDAC allows us to skip the heuristic choice for the training mixture of different domains or tasks.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \system{MDAC($\vlambda_1, \vlambda_2$)}&36.2&-&-&-&46.1&-&-&- \\
    \system{DDS($\vlambda_1, \vlambda_2$)}&36.5&-&-&-&44.4&-&-&- \\
    \system{MDAC($\vlambda_1, \vlambda_2$)}&-&57.2&53.8&-&-&-&-&- \\
    \system{DDS($\vlambda_1, \vlambda_2$)}&-&56.1&52.6&-&-&-&-&- \\
    \system{MDAC($\vlambda_1, \vlambda_2$)}&-&-&-&32&-&90&-&- \\
    \system{DDS($\vlambda_1, \vlambda_2$)}&-&-&-&32.9&-&91.1&-&- \\ \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, all the non empty columns \emph{correspond to the same bi-domain system}}
  \label{tab:bi-da}
\end{table*}

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*9{r|}} \hline
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean} & \multicolumn{1}{|c|}{\domain{news}} \\ \hline \hline
    \multicolumn{9}{|c|}{\sl Unsupervised domain adaptation} \\ \hline
    \system{Mixed-0}      & - & - & - & - & - & - & - & 22.4 \\
    \system{Mixed-0.25} & - & - & - & - & - & - & - & 22.7\\
    \system{Mixed-0.5}   & - & - & - & - & - & - & - & 23.6\\
    \system{Mixed-0.75} & - & - & - & - & - & - & - & 23.7\\
    \system{Mixed-1} & - & - & - & - & - & - & - & 23.5\\
    \hline \hline
    \system{MDAC($\vlambda_0, \vlambda_{news}$)} &-&-&-&-&-&-&-&23.1 \\
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &-&-&-&-&-&-&-&23.3 \\ \hline\hline 
    \multicolumn{9}{|c|}{\sl 30 clusters} \\ \hline
    \system{MDAC($\vlambda^*, \vlambda_d$)}&36.3&56.9&51.3&33.2&43.4&89.4&51.8&-\\
     \system{DDS($\vlambda^*, \vlambda_d$)}&35.3&55.2&49.4&30.6&43.3&89.9&50.6&-\\ \hline
  \end{tabular}
  \caption{Non empty columns correspond to the target domains. For a given line, each non empty column correspond to one expert system.}
  \label{tab:unsupervised-da}
\end{table*}
